---
title: "Increased financial dissatisfaction and worsening economic well-being in the United States (2018-2021)"
subtitle: ""
author: 
  - Roy Chan
thanks: "Code and data are available at: LINK."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(here)
library(tidyverse)
```

# Introduction
Throughout the past few years, we have seen numerous major economic events that have greatly influenced economic well-being. Supply constraints have created shortages for many consumer goods, driving up demand, increasing cost of living and thereby raising rates of inflation. Stock market crashes and industry shut downs lead to high levels of unemployment during most of 2020 which is still recovering to this day. Many if not all of these events can be attributed in part or fully to the COVID-19 pandemic, which has majorly disrupted global circumstances. Within the United States, there has been a growing movement commonly known as the 'Great Resignation', characterized by high levels of resignations and voluntary unemployment of workers. Worker dissatisfaction has been growing within the U.S. in part due to salary stagnation and perceived income inequality [@citePalladino], with the pandemic acting as a catalyst for the birth of the movement. 

Since 1972, the United States General Social Survey (US GSS) has been collecting data on opinions within the national population of America. Notably, economic well-being has been a consistent topic of interest within the survey, and as such we would like to explore whether the sentiments on topics such as satisfaction of financial situation and standard of living has changed in the most recent round of the survey, as compared to the one prior.

This paper was produced using `knitr` [@citeKnitr] and `bookdown` [@citeBookdown].

# Data
The survey data that is presented within this paper is obtained from the 2018 and 2021 rounds of the US GSS. The data, along with its corresponding codebooks, was retrieved from the GSS web portal [@citePortal]. All analysis was performed in `R` [@citeR], with the data being read and prepared using `haven` [@citeHaven], `here` [@citeHere], and `dplyr` [@citeDplyr]. Graphs were generated using `ggplot2` [@citeGgplot].

The US GSS was created to collect long term public opinion and sentiment within the United States. The main component of the survey is the replicating core, a central set of questions repeated in each year the survey was conducted for the purpose of exploring trends in opinion over time [@citeCC2021]. Additional modules present but not always consistently asked from year to year include topics relevant for the year. Examples of this for 2021 included but were not limited to: attitudes towards policing, racial discrimination, and opinions relating to trust in institutions and the government for handling the COVID-19 pandemic. A major strength of the US GSS is in its long existence and replicating core, allowing tracking of contemporary sentiments over a significant period of time. Prior to 2021, surveys were administered in person in an interview format by an interviewer, which as noted by the 2021 codebook resulted in high response rates of above 50%. 

For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core. Of interest are the following topics and corresponding questions:

Changes in financial situation - "During the last few years, has your financial situation been getting better, worse, or has it stayed the same?"

Opinion on family income - "Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?"

Standard of living will improve - "The way things are in America, people like me and my family have a good chance of improving our standard of living -- do you agree or disagree?"

Satisfaction with financial situation - "We are interested in how people are getting along financially these days. So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?"

## Survey methodology
The target population of the GSS are adults over the age of 18 who lived in non-institutional housing at the time of taking the survey. The intention is to make the obtained sample an accurate representation of the national population of the United States, thus a stratified sampling method was chosen. Since 2004, the US GSS has used the sampling frame adapted from the list of postal addresses maintained by the United States Postal Service. Areas are categorised with respect to their population densities as either urban or rural.

However, a significant limitation as discussed in the cumulative codebook [@citeCC] it is more often the case that urban and rural areas are interspersed. This causes difficulty for interviewers to effectively pursue interviewees, due to differences in density and ability to locate households in rural areas. From this, a multi-stage cluster sampling approach is applied. Firstly, stratification is performed- the entire sampling frame is grouped consisting of urban areas and one or more of its surrounding rural areas from which about half is selected. Secondly, each chosen area is divided into two segments according to whether they are part of a census tract or a block group. This allows for generating a high quality list of clustered addresses for interviewers to efficiently conduct the survey, reducing cost. 

As of 2004, non-response of the survey has been addressed through sub-sampling. For all of the households in the sample that were non-respondents a proportion, typically 50%, of those who were labelled as temporary non-respondents were pursued again. As such, those responses were weighted at 2 for the sub-sample to represent the entirety of the temporary non-respondents.

### Methodology changes to 2021 GSS
Owing to the COVID-19 pandemic, it was not possible to retain the physical interview format as used in all other years. Thus, the 2021 edition was primarily conducted through the web instead, with an alternative phone interview option. Participants were recruited through mail, which invited them to complete the survey online. Given that this method was utilised, the consideration of an interviewer having to physically perform interviews was no longer required. Thus, a stratified non-clustered sampling approach was used for a greater sample size. Another significant change was that non-respondents was not handled through sub-sampling. Instead, as specified by the 2021 codebook, adjustment in response weights in accordance with known population totals for each category the response belonged to was performed iteratively. 

# Results
Of the four chosen topics which we have focused on within this paper, all have displayed some level of shift in opinion between the years of 2018 and 2021. Generally, it appears economic well-being has decreased in 2021 overall, with fewer positive sentiments being recorded in most topics.

For the topic of changes in financial situation (Figure \@ref(fig:finalter)), all groups had saw a decrease in the response of 'Stayed the same', and reflected in increases for both positive and negative responses. The 18-29 age group saw the most significant change in any of the responses. Of all age groups, in 2018 they had the fewest percentage which considered their financial situation to be getting worse, to becoming the most in 2021. For ages 40-49 and 64-89, they saw an increase in the opinion of 'Getting better'; 50-64s almost no change between years (38.69% for 2018, 38.93% for 2021).

Perceptions of relative family income (Figure \@ref(fig:finrela)) also saw numerous changes in responses. It should be noted that as a result of the question phrasing, it is heavily reliant on subject perception on what they considered as the average income for an American family. Nonetheless, it is shown that all groups became more opinionated in 2021 by the decrease in 'average' responses. A greater portion of every group considered their income to be far above and above average in 2021, and only the groups 18-29 and 64-89 had increases in 'below average', while 'far below average' responses increased for 18-29, 30-39 and 40-49. 

With regards to the belief that standards of living will improve under the current circumstances in the United States (Figure \@ref(fig:goodlife)), previously all groups maintained a positive sentiment in 2018 which was not reflected in 2021. All groups had reduced agreement, which shifted to neutral to strongly disagreeing.

Lastly, satisfaction with personal and family income (Figure \@ref(fig:satfin)) became more neutral for all groups except 30-39, 40-49. These groups saw slight increases (<5%) in positive responses- however it was coupled alongside greater increase in negative ones. Similarly to changes in financial situations, 18-29 saw the largest decrease in positive satisfaction (-9.72%), with some becoming neutral (+2.25%) and more negative (+7.24%).

```{r import_data}
gss_data <- read_csv(here::here("inputs/data/gss_data.csv"))
```

```{r finalter, fig.align="center", fig.height = 6, fig.width = 5, fig.cap = 'Percentage of age group sharing opinion of changes in financial situation'}
# get proportion that each response represents, for corresponding age group and year
finalter <- gss_data %>% 
  drop_na() %>% 
  count(age, year, finalter) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
finalter <- finalter %>% 
  mutate(finalter = factor(finalter, levels = c('Getting better', 'Stayed the same', 'Getting worse')))

# plot the percentages, facet_wrap to separate plots by opinion
finalter %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(2018, 2021)) +
  facet_wrap(~finalter, scales = 'free_y', ncol = 1) +
  labs(title = 'Changes in financial situation over past few years', x = 'Year', y = '(%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```

```{r finrela, fig.align="center", fig.height = 7, fig.width = 5, fig.cap = 'Percentage of age group sharing perception of relative family income'}
# get proportion that each response represents, for corresponding age group and year
finrela <- gss_data %>% 
  drop_na() %>% 
  count(age, year, finrela) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
finrela <- finrela %>% 
  mutate(finrela = factor(finrela, levels = c('Far above average', 
                                              'Above average', 
                                              'Average',
                                              'Below average',
                                              'Far below average')))

# plot the percentages, facet_wrap to separate plots by opinion
finrela %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(2018, 2021)) +
  facet_wrap(~finrela, scales = 'free_y', ncol = 1) +
  labs(title = 'Perception of family income relative to others', x = 'Year', y = '(%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```

```{r goodlife, fig.align="center", fig.height = 7, fig.width = 5,fig.cap = 'Percentage of age group sharing belief that standard of living is likely to improve'}
# get proportion that each response represents, for corresponding age group and year
goodlife <- gss_data %>% 
  drop_na() %>% 
  count(age, year, goodlife) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
goodlife <- goodlife %>% 
  mutate(goodlife = factor(goodlife, levels = c('Strongly agree', 
                                              'Agree', 
                                              'Neither agree nor disagree',
                                              'Disagree',
                                              'Strongly disagree')))

# plot the percentages, facet_wrap to separate plots by opinion
goodlife %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(2018, 2021)) +
  facet_wrap(~goodlife, scales = 'free_y', ncol = 1) +
  labs(title = 'Standard of living is likely to improve...', subtitle = '...given current circumstances in the U.S.', x = 'Year', y = '(%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```

```{r satfin, fig.align="center", fig.height = 6, fig.width = 5, fig.cap = 'Percentage of age group sharing opinion of satisfaction with financial situation'}
# get proportion that each response represents, for corresponding age group and year
satfin <- gss_data %>% 
  drop_na() %>% 
  count(age, year, satfin) %>%
  group_by(age, year) %>%
  mutate(pct = (n / sum(n)) * 100)

# set levels of variable, so that it is ordered from best -> worst in facet_wrap
satfin <- satfin %>% 
  mutate(satfin = factor(satfin, levels = c('Pretty well satisfied', 'More or less satisfied', 'Not satisfied')))

# plot the percentages, facet_wrap to separate plots by opinion
satfin %>% ggplot(aes(x = year, y = pct, colour = age)) + 
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(2018, 2021)) +
  facet_wrap(~satfin, scales = 'free_y', ncol = 1) +
  labs(title = "Satisfaction of personal and family's financial situation", x = 'Year', y = '(%)', color = 'Age group') + 
  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
  theme_linedraw() + 
  scale_colour_brewer(palette = "Set1")
```
\newpage

# Discussion

## Growing work dissatisfaction and possible increase of income expectation for those aged 18-29
There is an interesting observation that can be made for respondents aged 18-29 that despite perceiving their financial situation as worsening in 2021, reinforced by a similar decrease in satisfaction, there has still been an increase in those who believe their income as being above average. We hypothesize that this seemingly contradictory combination of opinions could be the result of increasing income expectations and overall job dissatisfaction. Within the period of when the 2021 GSS had been conducted was the growing trend of the "Great Resignation", an economic movement of consisting of large numbers of voluntary resignations. In the United States, this was evidenced by worker quit rates rising drastically [@citeJOLTS] beginning in the middle of 2021, and has remained consistently high since. The trend has been theorized as being caused by a number of different factors, many of which have been catalyzed by the COVID-19 pandemic. Changes in mindsets and priorities such as establishing a healthier work-life balance and preference for remote work are common reasons for quitting [@citeHsu]. Of all workers, the trend appears to be most prevalent in younger ages [@citeSmart], which aligns with the opinions we have seen within the US GSS.

## Unanimous decreased belief of improving standard of living
In a similar sense, re-evaluating priorities in life can also be the reason for the decreased agreement for an improving standard of living during 2021- it could be the case that the standard of living has increased due to the shift in attitudes away from a work-centric lifestyle. What is notable about this topic in particular is in its unanimous change in opinion, indicating that situational changes in the U.S. has been felt across all age groups. Although not specified, it is probable that pandemic related changes are the most pronounced reason for this, causing the most significant life alterations to the national population. At the time of writing this paper many of the previous pandemic-related restrictions have been lifted, but at the same time major world events have occurred; it would be valuable for this topic to be explored again in the next round of the GSS to see whether this opinion holds or improves.

## Perception of income inequality
Opinions of relative financial situation in 2021 differed significantly from that of 2018, with no age group having more than 43% considering their income to be average, as compared to the lowest in 2018 of 40%. The decreased prominence of the average response has consequently increased that of both above and below average, possibly demonstrating a larger divide in income equality. The data for this topic is particularly intriguing as it also reflects perceived changes in income for "American families in general". Once again, this can be correlated with the events of the pandemic, disproportionately affecting those with higher living costs such as large families [@citeUnicef] and is also dependent on employment sector, ranging from little to no changes in working conditions to furloughs or permanent layoffs.

## Weaknesses
Several weaknesses of this paper include the utilisation of only two rounds of the US GSS in our analysis- it would have been beneficial to leverage this significant advantage of the survey and explore trends over a greater period of time. For the findings of change in opinion between 2018 and 2021, it is important to note the difference in survey methodology could have influenced responses. Higher than normal non-responses were recorded, indicating methodology changes has altered data collection. Furthermore, there is a level of uncertainty due to the subjective wording in questions. For instance, changes in financial situation uses the wording "last few years". Particularly for our discussion around the effects of the pandemic, this makes it difficult to say with certainty whether it caused changes in opinion. On the other hand, cognitive biases are often present in assessing financial situations [@citeBansal], so there is potential for recency bias due to the weight and scale of changes due to the pandemic.

## Next steps
The 2021 GSS has differed in its respondent demographic as compared to prior years, potentially due to format changes resulting in technical limitations. It would be appropriate to weight responses to correct sampling bias, and/or devising a way to address non-responses. It is unfortunate that due to limits in survey length some questions in prior years' replicating core that would have been valuable to our analysis were not reproduced, namely job security and ease of finding a similar job. As such, we aim to address this by including the questions in our supplemental survey.

\newpage

\appendix

# Supplementary survey


\newpage


# References


